---
title: "Drought analysis for gcam talk"
output: html_notebook
---


```{r setup}
library(survival)        # standard R package for survival analysis
library(stayinalive)     # functions we developed for processing time series.
library(ggplot2)
library(ggthemes)
```

```{r params}
beta <- 0.25
year <- 2001:2100
```

Generate 100 years of data for 10 basins.  Temperature increases over the course of the century
according to Hector's RCP8.5 pathway.  Droughts are between 1 and 2 years long in this
simulation.  The $\beta$ value of `r beta` means that each degree of global temperature change should
produce about a `r round(100*(exp(beta)-1))`% increase in drought risk relative to the baseline.

```{r gethectortemps}
hcore <- hector::newcore(system.file('input', 'hector_rcp85.ini', package='hector'))
invisible(hector::run(hcore,max(year)))
temps <- hector::fetchvars(hcore, year, hector::GLOBAL_TEMP())
Tgyr <- temps$value
Tgmon <- rep(Tgyr, rep(12, length(Tgyr)))   # convert to monthly values
```
```{r tempplot}
ggplot(mapping=aes(x=year, y=Tgyr)) + geom_line(color='lightgrey', size=1.25) + ylab('Global temperature') +
    theme_solarized_2(light=FALSE)
```


```{r simdata}
nyear <- length(Tgyr)
nmonth <- 12*nyear
nbasin <- 20

set.seed(867-5309)
basindata <- gents(nbasin, nmonth, Tgmon, beta=beta)
```

Plot the pattern of drought in the basin data.
```{r droughtplot}
bd <- t(basindata)
colnames(bd) <- seq(1,nbasin)
bddf <- as.data.frame(bd)
bddf$month <- seq(1,nrow(bddf))
bddf2 <- tidyr::gather(bddf, key='basin', value='value', -month, convert=TRUE)
bddf2$drought <- bddf2$basin * bddf2$value
bddf2$year <- (bddf2$month-1) / 12 + 2001
ggplot(data=bddf2, aes(x=year, y=drought)) + geom_point(color='lightgrey') + ylim(c(1,nbasin+1)) + ylab('Ensemble run') +
    theme_solarized_2(light=FALSE)
```


The code below converts the basin time series to the event data frame required by the `coxph`
function.  Then we feed that to a simple Cox model.

```{r surv}
events <- tsmat2event(basindata, Tgyr, 1)
events$basinid <- as.factor(events$basinid)
cph <- coxph(Surv(tstart, tstop, drought)~Tg, data=events)
summary(cph)
```

How much data do we need?  Can we detect the change with fewer basins?
```{r smalldata}
basindata10 <- basindata[1:10, ]
events10 <- tsmat2event(basindata10, Tgyr, 1)
events10$basinid <- as.factor(events10$basinid)
cph10 <- coxph(Surv(tstart, tstop, drought)~Tg, data=events10)
summary(cph10)
```
Marginal.  The fit is just a tiny bit above one SE away from the right answer.
However, the statistics are not good.  The confidence intervals include no effect.

```{r survcomp}
betafit <- cph$coefficients['Tg']
bh <- basehaz(cph, centered=FALSE)
h0 <- bh$hazard
calcsurv <- function(tg) {
    psi <- exp(tg*betafit)
    s <- exp(-psi*h0)
    data.frame(temperature=tg, time=bh$time, S=s)
}
pltdata <- dplyr::bind_rows(lapply(c(0,1,2,5), calcsurv))
pltdata$temperature <- paste(pltdata$temperature, 'deg C')
pltdata$timeyr <- pltdata$time / 12
ggplot(data=pltdata, aes(x=timeyr, y=S, color=temperature)) + geom_line(size=1.1) + 
    theme_minimal() + scale_color_manual(values=c('black','blue','green','red'))
```

Do the analysis of temperatures in the rcp8.5 scenario for the weird science section

```{r weirdsci}
ncfiles <- list.files('~/wrk/spva-paper1-devel/data', 'rcp85', full.names = TRUE)
tempdata <- get_ncdf_pixel(ncfiles, varnames=list(latvar='lat_2', lonvar='lon_2'))
tempdata <- reformat_data(tempdata)
dplyr::count(tempdata, tgbin)
ggplot(data=tempdata, aes(x=tgbin, y=Temperature)) + geom_boxplot() + 
    theme_minimal() + xlab('Global temperature') + ylab('Grid temperature')
```

Try fitting a Cox PH model using the grid temperature as the "time" variable and 
global temperature as the covariate.  Because each grid temperature is associated
with a single global temperature, the covariate does not have to be regarded as
"time" varying.

```{r cphweird}
cphweird <- coxph(Surv(Temperature, cens)~Tg, data=tempdata)
summary(cphweird)
betaweird <- cphweird$coefficients['Tg']
bhweird <- basehaz(cphweird, centered=FALSE)
h0weird <- bhweird$hazard
weirdcdf <- function(tg) {
    psi <- exp(tg*betaweird)
    s <- exp(-psi*h0weird)
    data.frame(Tg=factor(tg), Temperature=bhweird$time, P=1-s)
}
weirdpltdata <- dplyr::bind_rows(lapply(c(0,1,2,4), weirdcdf))
ggplot(data=weirdpltdata, aes(x=Temperature, y=P, color=Tg)) + geom_line(size=1.1) + 
    theme_minimal() + scale_color_manual(values=c('black','blue','green','red'))
```
