---
title: "Survival model test"
output: html_notebook
---


```{r setup}
library(survival)        # standard R package for survival analysis
library(stayinalive)     # functions we developed for processing time series.
doParallel::registerDoParallel(cores=4)
```


Generate 100 years of data for 225 basins.  Temperature increases over the course of the century
roughly according to Hector's RCP8.5 pathway.  Droughts are between 1 and 2 years long in this
simulation.  The $\beta$ value of 0.5 means that each degree of global temperature change should
produce about a 65% increase in drought risk relative to the baseline.

```{r simdata}
nyear <- 100
nmonth <- 12*nyear
nbasin <- 225

Tgyr <- seq(0.75, 4.5, length.out=nyear)        # Approximately RCP8.5
Tgmon <- rep(Tgyr, rep(12, length(Tgyr)))

set.seed(867-5309)
basindata <- gents(nbasin, nmonth, Tgmon, beta=0.5)
```

The code below converts the basin time series to the event data frame required by the `coxph`
function.  Then we feed that to a simple Cox model.

```{r surv}

events <- tsmat2event(basindata, Tgyr, 1)
events$basinid <- as.factor(events$groupid)
cph <- coxph(Surv(tstart, tstop, drought)~Tg, data=events)
summary(cph)
```

In this call we compute a clustered survival model, with basins as the clustering variable.
The idea here is that events from the same basin might be more similar to each other than
they are to events in other basins.  This has no effect on the estimate of $\beta$ (_i.e.,_ it
will be the same as in the model above); however, there will be an additional estimate of the
standard error (called the "robust standard error") that takes this into account.  We didn't
actually include any basin effects in the simulated data, so the robust standard error shouldn't
be that much different to the regular estimate.

```{r survcluster}
cph_clus <- coxph(Surv(tstart, tstop, drought) ~ Tg + cluster(basinid), data=events)
summary(cph_clus)
```

In the next example, we include a frailty term.  In this formulation, not only are clustered
items more similar to each other than they are to other observations, but they also have 
slightly different intrinsic hazard rates.  The additional increment to the hazard rate (the
"frailty" of the cluster) is chosen from a Gamma distribution with parameters that are solved
for in the model fit.  I _think_ this is equivalent to a partial pooling model of the sort 
described in McElreath (and elsewhere, of course), but I haven't exactly confirmed that.

At any rate, we didn't include any basin fixed effects, so there shouldn't be much to see
here.

```{r survfrail}
cph_frail <- coxph(Surv(tstart, tstop, drought) ~ Tg + frailty(basinid), data=events)
summary(cph_frail)
```

We may also want to add interaction terms to allow for the possibility that different basins
have different responses to climate change.  If results from previous papers are any guide, 
this is almost certainly the case.  One possibility would be to use an interaction term, like
in the example below.  As before, since we didn't actually include any of these effects in 
our simulation, we shouldn't see much difference to our old results.

```{r survinter}
cph_inter <- coxph(Surv(tstart, tstop, drought) ~ Tg + basinid:Tg, data=events)
summary(cph_inter)
```
